 %!TEX program = latexmk
\documentclass[12pt,fleqn]{article}
\usepackage{setspace}
\usepackage{float}

\usepackage[english]{babel}
\PassOptionsToPackage{table,xcdraw}{xcolor}
\usepackage{xcolor}
% \usepackage[utf8x]{inputenc}

\usepackage[style=apa, backend=biber]{biblatex}
\addbibresource{references.bib}

% A few common packages
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{array}
\usepackage{graphicx}
\usepackage{relsize}
\usepackage[titletoc]{appendix}
\usepackage{caption}
% \usepackage{xurl}
\captionsetup[table]{skip=10pt plus 0.01pt}
\usepackage{subcaption}  % \begin{subfigure}...\end{subfigure} within figure
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{makecell}
\usepackage{booktabs}
% \usepackage{ltablex}
\usepackage[autostyle, english = american]{csquotes}


% Uncomment the following to attempt to enforce Type 1 or TrueType 
% fonts. ProQuest does not want the type 3 fonts used by default as
% of Dec. 2019 - see 
% https://support.proquest.com/articledetail?id=kA01W000000k9o2SAA . 
% If you are unable to embed fonts such as 'Zapf Dingbats' or 
% 'Symbol', try using raster images (.jpg or .png) instead of vector 
%images (.pdf or .eps).
% \usepackage[T1]{fontenc} 

% plainpages=false fixes the "duplicate ignored" error with page counters
% Set pdfborder to 0 0 0 to disable colored borders around PDF hyperlinks
\usepackage[plainpages=false,pdfborder={0 0 0}]{hyperref}

\begin{document}

\begin{titlepage}
\pagenumbering{alph}
\centering
\vspace*{2cm}
{\LARGE\bfseries Modeling Decanalization with Homeostatic Reinforcement Learning\par}
\vspace{2cm}
{\large
Joshua Clingo\textsuperscript{1,*} and Jeff Yoshimi\textsuperscript{1}\par
}
\vspace{0.5cm}
{\small
ORCID: Joshua Clingo \texttt{0009-0003-9377-1532}; Jeff Yoshimi \texttt{0000-0001-7327-5599}\par
}
\vspace{1cm}
{\normalsize
\textsuperscript{1}School of Social Sciences, Humanities and Arts, University of California, Merced\par
}
\vspace{1cm}
{\normalsize
\textsuperscript{*}Corresponding author: Joshua Clingo\\
Email: jclingo@ucmerced.edu\par
}
\vfill
\end{titlepage}

\pagenumbering{arabic}

\begin{abstract}
Living organisms continuously manage their homeostatic setpoints but sometimes fall into the ruts and grooves of an entrenched or canalized behavioral repertoire. This manifests in a wide variety of ways, from overconsumption of scarce resources to repetitive unhealthy behavior, in cases of depression or anxiety. Clinicians are increasingly employing interventions, from psychoactive drugs to targeted stimulus regimes, to transiently decanalize entrenched behaviors and recanalize them into more adaptive patterns. These metaphors are long-standing, but have not been formally studied in explicit computational models. We present the first such model to our knowledge, a computational model of canalization, decanalization, and recanalization in a homeostatic navigation task. We compared homeostatic reinforcement learning (HRL) with a negative feedback (NF) reward function under matched canalization and decanalization conditions. In this setting, HRL agents showed improved post-decanalization homeostatic performance, relative to matched no-decanalization controls, whereas NF agents did not. We offer the model as a formal instantiation of a candidate mechanism by which temporary destabilization can produce more optimal behaviors in a constrained environment.
\end{abstract}

\vspace{0.5em}
\noindent\textbf{Keywords:} canalization, decanalization, homeostatic reinforcement learning, foraging, psychedelics, altered states

\include{content}

\section*{Declarations}

\subsection*{Declaration of conflicting interests}
The authors declare that there is no conflict of interest.

\subsection*{Funding}
\begin{sloppypar}
The authors received no financial support for the research, authorship, and/or publication of this article.
\end{sloppypar}

\subsection*{Use of AI}
The authors used AI agents to assist with generating and improving analysis code. AI was not used to generate or draft the scientific content of the manuscript.

\subsection*{Ethics approval}
Not applicable. This study used only simulated agents in a simulated environment; no human participants, nonhuman animals, biological specimens, or clinical records were involved.

\subsection*{Data and code availability}
The Unreal Engine source code, evaluation data, and analysis materials used in this study are openly available at \url{https://github.com/JClingo/decanalization}.

\sloppy
\printbibliography

\begin{appendices}
    \include{appendix}
\end{appendices}

\end{document}